RNAseq aCRY processing

### 2024-05-16 0_Tutorial/1_data_processing_Tutorial.Rmd
Add files via upload

# 0. Prepare System
BiocManager::install()

## -R_libraries
```{r r_libraries}
library(stringr)
library(R.utils)
library(RColorBrewer)
library(sessioninfo)
library(data.table)
library(plyr)
library(tidyverse)
library(tximeta)
library(tximport)
library(curl)
library(AnnotationHub)
library(DESeq2)
library(EnhancedVolcano)
library(pheatmap)
library(writexl)
library(biomaRt)
library(ape)
library(kableExtra)
library(knitr)

@@

@@

## -R_folders
```{r r_folders}
ifelse(Sys.info()["sysname"]== "Linux",
       s <- "/mnt/s",
       s <- "S:")
dir <- paste(s,"AG/AG-Scholz-NGS/Daten/Simon/Chlamy_RNASeq_aCRY",sep="/")
list.files(dir) %>% head()
gitdir <- paste(dir,"git_Chlamy_RNAseq_aCRY",sep="/")
list.files(gitdir) %>% head()

fastqdir <- paste(dir,"fastq",sep="/")
list.files(fastqdir) %>% tail()
fastqdir2 <- paste(s,"AG/AG-Scholz-NGS/Daten/Simon/P3044/fastq/",sep="/")
list.files(fastqdir2) %>% head()

indexDir <- paste(s,"AG/AG-Scholz-NGS/Daten/Simon/linux-ngs/salmon/salmon_index_chlamy/chlamy_index_v6.1",sep="/")

fastaPath <- file.path(s,"AG/AG-Scholz-NGS/Daten/Simon/linux-ngs/salmon/salmon_index_chlamy/Phytozone_v6.1/CreinhardtiiCC_4532_707_v6.0.hardmasked.fa.gz")
head(readLines(fastaPath,n=10))

gtfPath <- file.path(s,"AG/AG-Scholz-NGS/Daten/Simon/linux-ngs/salmon/salmon_index_chlamy/Phytozone_v6.1/CreinhardtiiCC_4532_707_v6.1.gene.gff3.gz")
readLines(gtfPath,n=10)

quantdir <- paste(dir,"quants",sep="/")

@@

## -Linux folders
```{bash linux_folders, eval=FALSE}
fastqdir="/mnt/s/AG/AG-Scholz-NGS/Daten/Simon/Chlamy_RNASeq_aCRY/fastq" #aCRY;
fastqdir2="/mnt/s/AG/AG-Scholz-NGS/Daten/Simon/P3044/fastq/" #pCRY;

dir="/mnt/s/AG/AG-Scholz-NGS/Daten/Simon/Chlamy_RNASeq_aCRY";

quantdir="$dir/quants";

indexdir="/mnt/s/AG/AG-Scholz-NGS/Daten/Simon/linux-ngs/salmon/salmon_index_chlamy/chlamy_index_v6.1";

@@

```{bash download data, eval=FALSE}

@@

## 1c. Mapping
```{bash mapping, eval=FALSE}
mamba activate salmon;

fastqfiles=$(ls $fastqdir/*_1*.fq.gz);
fastqfiles2=$(ls $fastqdir2/*R1*.fastq.gz);

# fastqfiles_all="${fastqfiles} ${fastqfiles2}";
fastqfiles_all=$fastqfiles;

echo "${fastqfiles:0:100}";
echo "${fastqfiles2:0:100}";
echo "${fastqfiles_all:0:100}";

# files: Unknown_BU327-002T0001_1.fq.gz;
for fn in $fastqfiles_all;
do
bfn=$(echo "`basename ${fn}`");
samp=${bfn%"_1.fq.gz"};
R1=$fn;
R2=$(echo "$R1" | sed 's/_1/_2/');
echo "Processing Sample: $samp";
test -f $R1 && echo "--> File: $R1 exists"
test -f $R2 && echo "--> File: $R2 exists"
outquant=${quantdir}/${samp}_quant
echo $outquant
echo $indexdir

salmon quant -i $indexdir -l A \
  -1 $R1 \
  -2 $R2 \
  -p 25 --validateMappings --gcBias -o $outquant

@@

@@

quantdir <- paste(dir,"quants",sep="/")

f <- list.files(path = quantdir)
files <- as.factor(str_remove(f,pattern ="_quant"))
# short form
sample <- str_remove(files,pattern ="Unknown_")
sample
# condition
library(readxl)
Samplepath <- filePath(path=dir, file="sample_table.xlsx")
Samplefile <- read_xlsx(Samplepath)
samplename <- Samplefile$clientName
Samplefile$treatment
Samplefile$condition <- paste(Samplefile$genotype, Samplefile$treatment, sep="_") %>% factor(levels = c("WT_dark","acry_dark","WT_blue","acry_blue","WT_red","acry_red","WT_3HL","WT_16HL","cia5_16HL"))
Samplefile$treatment <- Samplefile$treatment %>% factor(levels = c("dark","blue","red","light","3HL","16HL"))
# Make table
sample.table <- data.frame(Samplefile[,-3])
sample.table %>% kable()

@@

m.table <- data.frame(sample.table,mappingrates)
```

@@

xx <- barplot(height=mappingrates, cex.names=0.5, names=samplename, horiz=T, las=1, xlim=c(0,100))
text(x = mappingrates, y = xx, label = mappingrates, pos = 4, cex = 0.5, col = "red")

@@

## 2c. Tximeta

```{r tximeta}
library(sessioninfo)
library(data.table)
library(plyr)
library(tidyverse)
library(tximeta)
library(tximport)
library(curl)

getwd()
example.quant <- paste(quantdir,list.files(quantdir)[1],"quant.sf",sep="/")

@@

# first file:
example.table <- read.table(example.quant, header=T)
head(example.table)
basename(example.quant)
dirname(example.quant)
basename(dirname(example.quant))

# generate file list & prepare variables
files={}
for (i in list.dirs(path = dirname(dirname(example.quant)), full.names = TRUE, recursive = FALSE)) {
files <- c(files,paste(i,"/quant.sf",sep=""))
# print(basename(i))
# print(head(read.table(files[length(files)], header=T)))
# print (files[length(files)])
}
head(files)

@@

# names as working sample names
# see 'm.table' at mapping rates
coldata <- data.frame(files, names = run, stringsAsFactors=FALSE)
coldata <- data.frame(coldata,sample.table)
coldata %>% head() %>% kable()

# load tximeta
# with linked Transcriptome
## Chlamy
file.exists(indexDir, fastaPath, gtfPath)

@@

colData(se) %>% head() %>% kable()
# meta infos sind da (genotype, treatment,...)

@@

rowRanges(se) %>% kable()
# genome info
seqinfo(se)
# ?
head(assays(se)[["counts"]]) %>% kable()
# counts. THE DATA

@@

TXNAME <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "TXNAME", keytype = "GENEID", multiVals="first"))
# select()' returned 1:many mapping between keys and columns = 1 gene has many transcripts...
TXNAME.list <- (mapIds(edb,keys = mcols(gse)$gene_id, column = "TXNAME", keytype = "GENEID", multiVals="list"))
head(TXNAME.list)
# info is already there
CDSID <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "CDSID", keytype = "GENEID", multiVals="first"))
head(CDSID)
# ?
CDSNAME <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "CDSNAME", keytype = "GENEID", multiVals="first"))
head(CDSNAME)
# same as gene_id
EXONNAME <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "EXONNAME", keytype = "GENEID", multiVals="first"))
head(EXONNAME)
TXTYPE <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "TXTYPE", keytype = "GENEID", multiVals="first"))
head(TXTYPE)
# no useful info... no gene Symbol!!

@@

getwd()
# save(gse,file=paste(dir,"tximeta.RDS", sep="/"))
# gse <- 1
# load(file=paste(dir,"tximeta.RDS", sep="/"))

@@

# more gene info from master_annotation_table
annopath <- file.path(s,"AG/AG-Scholz-NGS/Daten/Simon/linux-ngs/salmon/salmon_index_chlamy/Phytozone_v6.1/CreinhardtiiCC_4532_707_v6.1.master_annotation_table.tsv")

anno <- read.delim2(annopath, header = T, sep = "\t",)
head(anno)[1:5,1:9] %>% kable()

@@

save(gse,file=paste(dir,"tximeta.RDS",sep="/"))
```

@@

### Deseq2 acry
```{r dds_acry}
design <- ~condition
# design <- ~genotype+treatment+genotype:treatment

dds <- DESeqDataSet(gse_acry, design=design)
colData(dds) %>% head() %>% kable()

##########################################\n
# filter all rows with rowsum = 0 ####
par(mfrow=c(1,2))
hist(log(counts(dds)), breaks=100, ylim = c(0,11000), xlim = c(0,10))
hist(counts(dds), breaks=1000000, ylim = c(0,100000), xlim = c(0,10))
# -> dip at log(counts(dds)=2-3)
par(mfrow=c(1,1))

@@

colData(dds) %>% head() %>% kable()
# add metadata
# colData(dds)$strain <- factor(colData(dds)$strain)

>>>>

# 0. Prepare System
## Download RStudio & R
https://posit.co/download/rstudio-desktop/

##  -install BiocManager
```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.18")
```

## -update Packages
```{r}
BiocManager::install()
```

## -install new packages
```{r}
BiocManager::install("PCAtools")

## -R_libraries
```{r r_libraries}
library(stringr)
library(R.utils)
library(RColorBrewer)
library(sessioninfo)
library(data.table)
library(plyr)
library(tidyverse)
library(tximeta)
library(tximport)
library(curl)
library(AnnotationHub)
library(DESeq2) # Main RNA-Seq Package
library(EnhancedVolcano)
library(pheatmap)
library(writexl)
library(biomaRt)
library(ape)
library(kableExtra)
library(knitr)

@@

library(readxl)

library(sessioninfo)
library(data.table)
library(plyr)
library(tidyverse)
library(tximeta)
library(tximport)
library(curl)

@@

## -R_folders
```{r r_folders}

# Server
server <- "/Volumes/biopro3/ExpBP/Chlamyland/RNA_seq"
list.files(server)

# for test folder
dir <- "/Volumes/biopro3/ExpBP/Chlamyland/RNA_seq/test" # Working directory
quantdir <- "/Volumes/biopro3/ExpBP/Chlamyland/RNA_seq/test/quants" # quant is the output of the mapping
figuredir <- "/Volumes/biopro3/ExpBP/Chlamyland/RNA_seq/test/figures"

@@

## -Connect to Linux machine (Enno))
```{r, eval=FALSE}
# run in terminal (mac)
# run in cmd.exe (Windows)

ssh -Y student@141.20.64.235
# Password: StudentExpBP

# activate salmon
conda activate /home/student/conda_salmon

# select server
cd /ExpBP/Chlamyland/RNA_seq
cd test

## -Linux folders
```{bash linux_folders, eval=FALSE}

# fastqdir=
# fastqdir2=

# dir="

# quantdir="$dir/quants";

# indexdir="/mnt/s/AG/AG-Scholz-NGS/Daten/Simon/linux-ngs/salmon/salmon_index_chlamy/chlamy_index_v6.1";

@@

## -Download via command line in linux
```{bash download data, eval=FALSE}

@@

## Salmon Mapping
### Make index
```{bash make_index}

# Instructions
# https://salmon.readthedocs.io/en/latest/salmon.html

cd /ExpBP/Chlamyland/RNA_seq
cd test

salmon index -t gentrome_chlamy.v6.1.fa.gz -i Chlamy_index_Olga  -k 31 -p 10

# --decoys decoys_chlamy_v6.1.txt

### 1c. Mapping / Quantifying
```{bash mapping, eval=FALSE}

conda activate /home/student/conda_salmon
cd /ExpBP/Chlamyland/RNA_seq
cd test

# test folder
quantdir="/ExpBP/Chlamyland/RNA_seq/test/quants"
indexdir="/ExpBP/Chlamyland/RNA_seq/test/Chlamy_index_Olga"
fasta="/ExpBP/Chlamyland/RNA_seq/test/fastq"

# "Real" data Chlamy_RNASeq_aCRY
quantdir="/ExpBP/Chlamyland/RNA_seq/Chlamy_RNASeq_aCRY/quants_v2"
indexdir="/ExpBP/Chlamyland/RNA_seq/test/Chlamy_index_Olga"
fasta="/ExpBP/Chlamyland/RNA_seq/Chlamy_RNASeq_aCRY/fastq"

# go to fastq folder
fastqfiles=$(\ls *_1.fq.gz)

echo "${fastqfiles:0:100}";

cd $fasta
# files: Unknown_BU327-002T0001_1.fq.gz;
for fn in $fastqfiles;
do
bfn=$(echo "`basename ${fn}`");
dfn=$(echo "`dirname ${fn}`");

samp=$(echo "$bfn" | sed 's/_1.fq.gz//');
R1=$(echo $fasta"/"$bfn);
R2=$(echo "$R1" | sed 's/_1/_2/');
echo "Processing Sample: $samp";
echo $R1
echo $R2

outquant=${quantdir}/${samp}_quant
echo $outquant
echo $indexdir

salmon quant -i $indexdir -l A \
  -1 $R1 \
  -2 $R2 \
  -p 15 --validateMappings --gcBias -o $outquant

@@

# code end

# test if file is there
if [ -e $R1 ]
then
    echo "ok"
else
    echo "nok"
fi

@@

quantdir # defined above

f <- list.files(path = quantdir)
files <- as.factor(str_remove(f,pattern ="_quant"))

# short form
sample <- str_remove(files,pattern ="Unknown_")
sample

# condition
Samplepath <- filePath(path="/Users/olga/Desktop/RNA-Seq Simon/Chlamy/Server/aCRY/", file="sample_table.xlsx")
Samplefile <- read_xlsx(Samplepath)
samplename <- Samplefile$clientName
Samplefile$treatment
Samplefile$condition <- paste(Samplefile$genotype, Samplefile$treatment, sep="_") %>% factor(levels = c("WT_dark","acry_dark","WT_blue","acry_blue","WT_red","acry_red","WT_3HL","WT_16HL","cia5_16HL"))
Samplefile$treatment <- Samplefile$treatment %>% factor(levels = c("dark","blue","red","light","3HL","16HL"))
# Make table
sample.table <- data.frame(Samplefile[,-3])
sample.table %>% kable()

@@

m.table <- data.frame(sample.table,mappingrates) # here used earlier only for columns [1:2,], otherwise for the whole table use sample.table
```

@@

xx <- barplot(height=mappingrates, cex.names=0.5, names=samplename[2], horiz=T, las=1, xlim=c(0,100))
text(x = mappingrates, y = xx, label = mappingrates, pos = 4, cex = 0.5, col = "red")

@@

## 2c. Tximeta
combine counts from different transcripts (splice variants) to gene level

```{r tximeta}

example.quant <- paste(quantdir,list.files(quantdir)[1],"quant.sf",sep="/")

@@

# first file:
example.table <- read.table(example.quant, header=T)
head(example.table)
dim(example.table)
basename(example.quant)
dirname(example.quant)
basename(dirname(example.quant))

# generate file list & prepare variables
files={}
for (i in list.dirs(path = quantdir, full.names = TRUE, recursive = FALSE)) {
files <- c(files,paste(i,"/quant.sf",sep=""))
}
head(files)
length(files)

@@

# names as working sample names
# see 'm.table' at mapping rates
coldata <- data.frame(files, names = run, stringsAsFactors=FALSE)
coldata <- data.frame(coldata,sample.table[1:2,])
coldata %>% head() %>% kable()

# load tximeta
# with linked Transcriptome
## Chlamy

indexDir="/Volumes/biopro3/ExpBP/Chlamyland/RNA_seq/test/Chlamy_index_Olga" # salmon index
fastaPath="/Volumes/biopro3/ExpBP/Chlamyland/RNA_seq/salmon_index_chlamy/Phytozone_v6.1/CreinhardtiiCC_4532_707_v6.0.hardmasked.fa.gz" # genomic fasta
gtfPath="/Volumes/biopro3/ExpBP/Chlamyland/RNA_seq/salmon_index_chlamy/Phytozone_v6.1/CreinhardtiiCC_4532_707_v6.1.gene.gff3.gz"

file.exists(indexDir, fastaPath, gtfPath)

@@

# Column infos = Sample infos -> colData(se)
colData(se) %>% head() %>% kable()

# Row Infos = Gene Infos -> mcols(se)
mcols(se) %>% head() %>% kable()

@@

# Gene length
rowRanges(se) %>% head() %>% kable()

head(assays(se)[["counts"]]) %>% kable()
# counts. THE DATA

@@

TXNAME <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, keytype = "GENEID", column = "TXNAME", multiVals="first"))
# select()' returned 1:many mapping between keys and columns = 1 gene has many transcripts...
TXNAME.list <- (mapIds(edb,keys = mcols(gse)$gene_id, column = "TXNAME", keytype = "GENEID", multiVals="list"))
head(TXNAME.list)
# info is already there
CDSID <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "CDSID", keytype = "GENEID", multiVals="first"))
head(CDSID)
# ?
CDSNAME <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "CDSNAME", keytype = "GENEID", multiVals="first"))
head(CDSNAME)
# same as gene_id
EXONNAME <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "EXONNAME", keytype = "GENEID", multiVals="first"))
head(EXONNAME)
TXTYPE <- as.character(mapIds(edb,keys = mcols(gse)$gene_id, column = "TXTYPE", keytype = "GENEID", multiVals="first"))
head(TXTYPE)
# no useful info... no gene Symbol!!

@@

getwd()
save(gse,file=paste(dir,"tximeta_test.RDS", sep="/"))
# gse <- 1
# load(file=paste(dir,"tximeta.RDS", sep="/"))

@@

# more gene info from master_annotation_table
annopath <- file.path("/Volumes/biopro3/ExpBP/Chlamyland/RNA_seq/salmon_index_chlamy/Phytozone_v6.1/CreinhardtiiCC_4532_707_v6.1.master_annotation_table.tsv")

anno <- read.delim2(annopath, header = T, sep = "\t",)
head(anno)[1:5,] %>% kable()

@@

save(gse,file=paste(dir,"tximeta_test.RDS",sep="/"))
```

@@

### Deseq2 acry
```{r dds_acry}
design <- ~condition
# design <- ~genotype+treatment+genotype:treatment
colData(gse)
dds <- DESeqDataSet(gse, design=~ 1)
dds <- DESeqDataSet(gse, design=design)

##########################################\n
# filter all rows with rowsum = 0 ####
par(mfrow=c(1,2))
hist(log(counts(dds)), breaks=100)
hist(counts(dds), breaks=1000000)
# -> dip at log(counts(dds)=2-3)
par(mfrow=c(1,1))

@@

# add metadata
# colData(dds)$strain <- factor(colData(dds)$strain)
